Using SMT engine to generate Symbolic Automata\thanks{This work was partially funded by the Associated Team FM4CPS between INRIA and ECNU, Shanghai}

### 2018-01-22 2017-Xudong-SMT/FM-2018-extended.tex
pour soumissino temporaire

\newcommand{\TODO}[1]{\textcolor{red}{\textbf{[TODO:#1]}}}
\newcommand{\NOTE}[1]{\textcolor{blue}{\textbf{[NOTE:#1]}}}
\newcommand{\ERIC}[1]{\textcolor{blue}{#1}}
\newcommand{\OTvar}{\texttt}
\newcommand{\OTland}{\;\land\ }
\definecolor{airforceblue}{rgb}{0.26, 0.44, 0.56}
\newcommand{\QIN}[1]{\textcolor{airforceblue}{#1}}
\newcommand{\coloncolon}{{:\hspace{-.2ex}:}}
\makeatletter
\newcommand{\raisemath}[1]{\mathpalette{\raisem@th{#1}}}
\newcommand{\raisem@th}[3]{\raisebox{#1}{$#2#3$}}
\makeatother

@@

\begin{algorithm}[h]
  \caption{Open Automaton Generation}
  \label{alg1}
\begin{algorithmic}[1]
\Require The pNet \hl{$P$} (cannot be a hole)
        \qquad \hl{\textbackslash\textbackslash~SB: was `` pn ''}
\State Initialize sets $U=\{\mathit{InitState}(pn)\}$ and $E=\emptyset$,
for unexplored and explored global states, respectively; $L=\emptyset$ for the resulting OTs;
\While{!$\mathit{isEmpty}(U)$}
	\State Choose $S$ in \hl{$U\setminus E$}; add $S$ in $E$;
        \qquad \hl{\textbackslash\textbackslash~SB: was `` Choose $S$ in $U$ ''}
	\State $\mathit{OT} = \mathit{MakeTransitions}(P, S)$;
        \State Store $\mathit{OT}$ in $L$; 
	\For {each $\mathit{OT} \in L$}
        \State Let $T$ be the target global state of $\mathit{OT}$
	\IfThen{($T \not\in U$ \&\& $T \not\in E$)}{Add $T$ into $U$;}
	\EndFor
\EndWhile
\State Prune the unsatisfiable open transitions in $L$ using the SMT solver;
\State \Return $\mathit{OA}=\mathit{Reachable}(\mathit{InitState}(pn),L)$;

@@

The inside loop (\emph{MakeTransitions} method) applies recursively
the semantic rules following the structure of the pNet.
When applied to a pLTSs at the leaves, we simply take the pLTS transitions of the
corresponding local state and use the semantic rule (see TR1 in
App.~\ref{appendix:semRules}) to build the OT.\footnote{%
%
  \hl{We omit detailed presentation of this case for the sake of brevity.}
%
}
When applied to a pNet node we use two methods, combining and
matching, to generate the open transitions in a hierarchical manner,
as shown in Alg.~\ref{alg2}. This method directly manages the
holes of the node, so \emph{MakeTransitions} is never called on a hole.

@@

In this paper, we presented an algorithm for generating an open
automaton representing the semantics of an open pNet.
We have also presented our implementation, which comprises two main
parts.
\noteEM{doit pouvoir se reecrire de maniere plus simple et concise...}
\hl{Firstly, we
compute} all the possible open transitions.
The algorithm combines the actions of subnets and
hole behaviors, which are then matched with the
synchronization vectors by the generated
predicate.  Some of the open transitions obtained at this
intermediate step, constructed in a structural manner from all
possible combinations of possible logical predicates, may
contain predicates which do not represent any possible concrete
instantiations. 
Secondly, to get rid of these useless transitions, we use the SMT solver Z3 for checking the
satisfiability of the predicate in each open transition. 
To this end, we encode into Z3 the representations of the action algebra and
the predicates before submitting them to the
Z3 solver. 
%
%\hl{%
%
In this paper, we used a running example, based on a BIP
architecture from an earlier nanosatellite case
study~\cite{CubETH-case-study}.  This example shows that
open-automata-based semantics can be instrumental in verifying the
properties enforced by the architectures through an encoding into open
pNets.  This encoding\mdash which we intend to formalise and prove
correct in a separate paper\mdash also opens the way for an extension
of BIP architectures with the transfer of data among variables of
different components.  Indeed, such data transfer can be easily
encoded using the predicates associated to synchronisation vectors in
open pNets.  The encoding of open transitions into SMTlib and the
availability of theories can guide the definition of such an extension.
%
%}
%
Our case-studies show that our encoding
successfully identifies the unsatisfiable open transitions and that
the resulting automata correctly reflect the expected 
movements of the encoded process expressions.

@@

%% \todoMargin{Change perspective: we are working on the formal extension of
%%   BIP architectures, and their semantics in terms of pNets. Then
%%   bisimulation and model-checking principles adapted to open pNet
%%   systems are under development}
%
Naturally, our next goals after the generation of the open automata
will be to model-check logical properties, and to check equivalence of pNets. While
model-checking open automata (with a suitable logic including
predicates on data) seems easy to define, equivalence checking is more challenging.
In a previous paper~\hl{\cite{x}}, we
have already found the FH-bisimulation, inspired by~\cite{deSimone85}, to be a
suitable definition. But weak equivalences, or refinements, will
definitely be useful when comparing different pNets with
different structure. For bisimulation, we foresee that SMT methods
will be the basis for comparison of open transitions. 
%However, the usual approach of computing bisimulation by partition
%% refinement does not work easily with symbolic transitions. In a first
%% step we may want instead to define and implement an algorithm for
%% checking that a given equivalence relation between open automata 
%% states is indeed a FH-Bisimulation. Finding automatically a
%% bisimulation relation or the most general bisimulation will be more
%% challenging.
%% It appears also that when comparing different pNets with
%% different structure, strong bisimulation will be too strong a notion,
%% and that we will have to define proper notions of weak equivalence or
%% refinements.

\paragraph{Scaling up.} One important motivation of this work is to
attack the complexity of verification of realistic systems by a
composition and parametric approach. Still one may wonder if the price
for analysing our our symbolic transitions will not make the approach
innaplicable. As a first exercice in this direction, we tried a
slighly bigger example, assembling 2 Failure controlers. We give some
figures from this try in appendix~ref{}, showing that Z3 can check the
satisfiablility of a 90K open transitions in a metter of minutes.

>>>>

%\newcommand{\TODO}[1]{\textcolor{red}{\textbf{[TODO:#1]}}}
\newcommand{\TODO}[1]{}
\newcommand{\NOTE}[1]{\textcolor{blue}{\textbf{[NOTE:#1]}}}
\newcommand{\ERIC}[1]{\textcolor{blue}{#1}}
\newcommand{\OTvar}{\texttt}
\newcommand{\OTland}{\;\land\ }
\definecolor{airforceblue}{rgb}{0.26, 0.44, 0.56}
\newcommand{\QIN}[1]{\textcolor{airforceblue}{#1}}
\newcommand{\coloncolon}{{:\hspace{-.2ex}:}}
\makeatletter
\newcommand{\raisemath}[1]{\mathpalette{\raisem@th{#1}}}
\newcommand{\raisem@th}[3]{\raisebox{#1}{$#2#3$}}
\makeatother

@@

\begin{algorithm}[h]
  \caption{Open Automaton Generation}
  \label{alg1}
\begin{algorithmic}[1]
\Require A pNet $P$ (cannot be a hole)
        %\qquad \hl{\textbackslash\textbackslash~SB: was `` pn ''}
\State Initialize sets $U=\{\mathit{InitState}(P)\}$ and $E=\emptyset$,
for unexplored and explored global states, respectively; $L=\emptyset$ for the resulting OTs;
\While{!$\mathit{isEmpty}(U)$}
	\State Choose $S$ in $U\setminus E$; add $S$ in $E$;
        %\qquad \hl{\textbackslash\textbackslash~SB: was `` Choose $S$ in $U$ ''}
	\State $\mathit{OT} = \mathit{MakeTransitions}(P, S)$;
        \State Store $\mathit{OT}$ in $L$; 
	\For {each $\mathit{OT} \in L$}
        \State Let $T$ be the target global state of $\mathit{OT}$
	\IfThen{($T \not\in U$ \&\& $T \not\in E$)}{Add $T$ into $U$;}
	\EndFor
\EndWhile
\State Prune the unsatisfiable open transitions in $L$ using the SMT solver;
\State \Return $\mathit{OA}=\mathit{Reachable}(\mathit{InitState}(P),L)$;

@@

The inside loop (\emph{MakeTransitions} method) applies recursively
the semantic rules following the structure of the pNet.
When applied to a pLTSs at the leaves, we simply take the pLTS transitions of the
corresponding local state and use the semantic rule (see TR1 in
App.~\ref{appendix:semRules}) to build the OT.\footnote{%
%
  We omit detailed presentation of this case for the sake of brevity.
%
}
When applied to a pNet node we use two methods, combining and
matching, to generate the open transitions in a hierarchical manner,
as shown in Alg.~\ref{alg2}. This method directly manages the
holes of the node, so \emph{MakeTransitions} is never called on a hole.

@@

In this paper, we presented an algorithm for generating an open
automaton representing the semantics of an open pNet.
We have also presented our implementation, which comprises two main
parts.
%\noteEM{doit pouvoir se reecrire de maniere plus simple et concise...}
%\hl{
  Firstly, we
  compute
%}
all the possible open transitions.
The algorithm combines the actions of subnets and
hole behaviors, which are then matched with the
synchronization vectors by the generated
predicate.  Some of the open transitions obtained at this
intermediate step, constructed in a structural manner from all
possible combinations of possible logical predicates, may
contain predicates which do not represent any possible concrete
instantiations. 
Secondly, to get rid of these useless transitions, we use the SMT solver Z3 for checking the
satisfiability of the predicate in each open transition. 
To this end, we encode into Z3 the representations of the action algebra and
the predicates before submitting them to the
Z3 solver. 
%
%\hl{%
%
In this paper, we used a running example, based on a BIP
architecture from an earlier nanosatellite case
study~\cite{CubETH-case-study}.  This example shows that
open-automata-based semantics can be instrumental in verifying the
properties enforced by the architectures through an encoding into open
pNets.  This encoding\mdash which we intend to formalise and prove
correct in a separate paper\mdash also opens the way for an extension
of BIP architectures with the transfer of data among variables of
different components.  Indeed, such data transfer can be easily
encoded using the predicates associated to synchronisation vectors in
open pNets.  The encoding of open transitions into SMTlib and the
availability of theories can guide the definition of such an extension.
%
%}
%
Our case-studies show that our encoding
successfully identifies the unsatisfiable open transitions and that
the resulting automata correctly reflect the expected 
movements of the encoded process expressions.

@@

%% \todoMargin{Change perspective: we are working on the formal extension of
%%   BIP architectures, and their semantics in terms of pNets. Then
%%   bisimulation and model-checking principles adapted to open pNet
%%   systems are under development}
%
Naturally, our next goals after the generation of the open automata
will be to model-check logical properties, and to check equivalence of pNets. While
model-checking open automata (with a suitable logic including
predicates on data) seems easy to define, equivalence checking is more challenging.
In~\cite{henrio:Forte2016}, we
have already found the FH-bisimulation, inspired by~\cite{deSimone85}, to be a
suitable definition. But weak equivalences, or refinements, will
definitely be useful when comparing different pNets with
different structure. For bisimulation, we foresee that SMT methods
will be the basis for comparison of open transitions. 
%However, the usual approach of computing bisimulation by partition
%% refinement does not work easily with symbolic transitions. In a first
%% step we may want instead to define and implement an algorithm for
%% checking that a given equivalence relation between open automata 
%% states is indeed a FH-Bisimulation. Finding automatically a
%% bisimulation relation or the most general bisimulation will be more
%% challenging.
%% It appears also that when comparing different pNets with
%% different structure, strong bisimulation will be too strong a notion,
%% and that we will have to define proper notions of weak equivalence or
%% refinements.

\emph{Scaling up.} One important motivation of this work is to
attack the complexity of verification of realistic systems by a
composition and parametric approach. Still one may wonder if the price
for analysing our our symbolic transitions will not make the approach
innaplicable. As a first exercice in this direction, we tried a
slighly bigger example, assembling 2 Failure controlers. We give some
figures from this try in appendix~ref{}, showing that Z3 can check the
satisfiablility of a 90K open transitions in a metter of minutes.
